Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-11 paper/paper.tex
DFAs->DFA

The specific contributions of this work are
  \begin{itemize}
  \item  We formulate  a  liveness  analysis for  a  lazy first  order
    functional language. We show how  to compute function summaries in
    a  context  independent  manner  to achievs  a  context  sensitive
    analysis. We  further describe the transformation  of the liveness
    equations to grammar rules in a  way that the language accepted by
    the resulting grammar describes live heap locations.\mycomment{the
      proof of its correctness}
  \item We prove  the undecidability of the grammar  that results from
    the  analysis.   To overcome  this,  we  present an  algorithm  to
    transform grammar  to finite  state automata  (DFA) such  that the
    language accepted by the  DFAs over-approximates the liveness (and
    hence a safe approximation).
  \item We have implemented a garbage  collector that can use the DFAs
    to  retain live  cells that  also include  closures. We  have done
    extensive  experiments with  this garbage  collector and  obtained
    empirical results  that show  the effectiveness  of liveness-based
    garbage collection.
  \end{itemize}

@@

We circumvent the problem  of undecidability by over- approximating the
CFG by non-deterministic finite state automata (NFA). The NFA are then
simplified using  the $\hookrightarrow$ rules.  Finally the simplified
NFAs are converted to DFAs.

%------------------------------------------------------------%
\subsubsection{From CFGs to GC-ready DFAs}
\label{sec:NFA-approx}
 We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
 approximate a CFG to a {\em strongly regular\/} grammar.  For
 example, the grammar
 fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
 Mohri-Nederhof transformation is:
 \begin{eqnarray*}
   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
   \acdr\var{\Df{\length}{1}}
   \mid \clazy\var{\Df{\length}{1}}\\
   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
   \mid \epsilon
 \end{eqnarray*}

@@

\section{The Garbage Collection Scheme}
\label{sec:GC-scheme}
Our experimental setup consists of  an interpreter for our language, a
liveness  analyzer, and  a single  generation copying  collector.  The
garbage collector  can be  easily configured to  work on the  basis of
reachability (RGC  mode) or use  the liveness DFAs (LGC  mode).  While
exploring the  activation record of a  function, LGC uses  the DFAs at
the latest program  point traversed in the function's  body.  Thus, if
LGC is invoked  at the program point $\pi$, it  would use the liveness
DFAs  at  $\pi$ itself  for  exploring the  root  set  of the  current
activation record. For  any other activation record in  the stack, say
for a function ${\mathit f}$  calling ${\mathit g}$ in the call-chain,
the evaluation  point in ${\mathit f}$  which resulted in  the call to
${\mathit g}$ would be used.

@@

%% \comment{
%% While exploring the activation record of a function, LGC uses the DFAs
%% at  the latest  program  point traversed  in the function's body.
%% Thus, if  LGC
%% is invoked  at the program point   $\pi$,  it would  use  the
%% liveness  DFAs at  $\pi$ itself for exploring the root set of the
%% current activation record. For any other activation record in the
%% stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the
%% call-chain,
%% the DFAs at the point of call to  ${\mathit g}$ would be used.}

@@

%% To  avoid scalability  issues (and  to  avoid performing  part of  the
%% liveness computation at run time)  we defined an 0-CFA version of this
%% liveness analysis  in which  demands for function  $f$ at  all calling
%% contexts are conflated into  a single demand $\sigma_f$.  This enabled
%% us to treat the liveness equations symbolically obtaining context-free
%% grammars for liveness at each GC point (calls to user functions and to
%% $\CONS$).  These were then converted to DFAs for run-time consultation
%% by  the garbage collector.  Experiments confirm  the precision  of the
%% analysis.

@@

  Orthogonally,  we plan  to improve  the efficiency  of the  liveness
  based garbage collector using heuristics  such as limiting the depth
  of  DFAs,   merging  nearly-equivalent   states  and   using  better
  representation  and algorithms  for automata  manipulation. We  also
  need  to   investigate  the  interaction  of   liveness  with  other
  collection schemes, such as incremental and generational collection.
  In  summary, we  need to  investigate  ways to  make liveness  based
  garbage collection attractive for practical collectors.  }

>>>>

The specific contributions of this work are
  \begin{itemize}
  \item  We formulate  a  liveness  analysis for  a  lazy first  order
    functional language. We show how  to compute function summaries in
    a  context  independent  manner  to achievs  a  context  sensitive
    analysis. We  further describe the transformation  of the liveness
    equations to grammar rules in a  way that the language accepted by
    the resulting grammar describes live heap locations.\mycomment{the
      proof of its correctness}
  \item We prove  the undecidability of the grammar  that results from
    the  analysis.   To overcome  this,  we  present an  algorithm  to
    transform grammar  to finite  state automata  (DFA) such  that the
    language accepted by the  DFA over-approximates the liveness (and
    hence a safe approximation).
  \item We have implemented a garbage  collector that can use the DFA
    to  retain live  cells that  also include  closures. We  have done
    extensive  experiments with  this garbage  collector and  obtained
    empirical results  that show  the effectiveness  of liveness-based
    garbage collection.
  \end{itemize}

@@

We circumvent the problem  of undecidability by over- approximating the
CFG by non-deterministic finite state automata (NFA). The NFA are then
simplified using  the $\hookrightarrow$ rules.  Finally the simplified
NFAs are converted to DFA.

%------------------------------------------------------------%
\subsubsection{From CFGs to GC-ready DFA}
\label{sec:NFA-approx}
 We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
 approximate a CFG to a {\em strongly regular\/} grammar.  For
 example, the grammar
 fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
 Mohri-Nederhof transformation is:
 \begin{eqnarray*}
   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
   \acdr\var{\Df{\length}{1}}
   \mid \clazy\var{\Df{\length}{1}}\\
   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
   \mid \epsilon
 \end{eqnarray*}

@@

\section{The Garbage Collection Scheme}
\label{sec:GC-scheme}
Our experimental setup consists of  an interpreter for our language, a
liveness  analyzer, and  a single  generation copying  collector.  The
garbage collector  can be  easily configured to  work on the  basis of
reachability (RGC  mode) or use  the liveness DFA (LGC  mode).  While
exploring the  activation record of a  function, LGC uses  the DFA at
the latest program  point traversed in the function's  body.  Thus, if
LGC is invoked  at the program point $\pi$, it  would use the liveness
DFA  at  $\pi$ itself  for  exploring the  root  set  of the  current
activation record. For  any other activation record in  the stack, say
for a function ${\mathit f}$  calling ${\mathit g}$ in the call-chain,
the evaluation  point in ${\mathit f}$  which resulted in  the call to
${\mathit g}$ would be used.

@@

%% \comment{
%% While exploring the activation record of a function, LGC uses the DFA
%% at  the latest  program  point traversed  in the function's body.
%% Thus, if  LGC
%% is invoked  at the program point   $\pi$,  it would  use  the
%% liveness  DFA at  $\pi$ itself for exploring the root set of the
%% current activation record. For any other activation record in the
%% stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the
%% call-chain,
%% the DFA at the point of call to  ${\mathit g}$ would be used.}

@@

%% To  avoid scalability  issues (and  to  avoid performing  part of  the
%% liveness computation at run time)  we defined an 0-CFA version of this
%% liveness analysis  in which  demands for function  $f$ at  all calling
%% contexts are conflated into  a single demand $\sigma_f$.  This enabled
%% us to treat the liveness equations symbolically obtaining context-free
%% grammars for liveness at each GC point (calls to user functions and to
%% $\CONS$).  These were then converted to DFA for run-time consultation
%% by  the garbage collector.  Experiments confirm  the precision  of the
%% analysis.

@@

  Orthogonally,  we plan  to improve  the efficiency  of the  liveness
  based garbage collector using heuristics  such as limiting the depth
  of  DFA,   merging  nearly-equivalent   states  and   using  better
  representation  and algorithms  for automata  manipulation. We  also
  need  to   investigate  the  interaction  of   liveness  with  other
  collection schemes, such as incremental and generational collection.
  In  summary, we  need to  investigate  ways to  make liveness  based
  garbage collection attractive for practical collectors.  }
